---
title: 'Pacific gaint salamenders in Mack Creek: Size and weight vs. channel location in old growth and clear cut forest section'
author: "Shuhan Song"
date: "November 23, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# attach packages
library(tidyverse)
library(janitor)
library(here)
library(janitor)
library(ggbeeswarm)
library(kableExtra)
library(plotly)
library(effsize)
library(car)

# load data
salamander <- read_csv("mack_creek_vertebrates.csv") %>% 
  clean_names() %>% 
  select(year, species, section, unittype, weight) %>% 
  filter(species == "DITE") %>% 
  filter(unittype %in% c("P", "C", "SC"))
```

### Introduction

### Data and Methods

### Results

#### A. Visually compare annual salamander counts in old growth (OG) and clear cut (CC) sections of Mack Creek.

The annual count of Pacific gaint salamanders was visualized for comparison between two groups of forest sections (CC/OG) of Mack Creek from 1993-2017 (Figure 1). In both sections, the count of Pacific gaint salamanders generally increased from 1993 to 2011, then fluctuated from 2011 to 2017 but did not drop back to the lowest levels in late 1990s. In most years, salamanders in OG sections had greater counts and seemed to be more robust to survival pressures than those in CC sections.

```{r}
# count annual count of salamanders in sections (OG, CC)
annual_count <- salamander %>% 
  count(year, section) %>% 
  mutate(section = ifelse(section == "CC", "Clear cut", "Old growth")) # write out the full names for ggplotly()

annual_count_plot <- ggplot(data = annual_count, 
       aes(x = year, y = n)) +
  geom_point(aes(color = section)) +
  geom_line(aes(color = section)) +
  scale_x_continuous(breaks = c(1995, 2000, 2005, 2010, 2015)) +
  scale_y_continuous(limits = c(0, 400)) +
  scale_color_discrete(name = "Sections") +
  theme_light() +
  labs(x = "Year",
       y = "Count")

ggplotly(annual_count_plot)
```

_**Figure 1**: Annual count of Pacific gaint salamanders in clear cut (CC) or old growth (OG) forest sections of Mack Creek (1993-2017). The red line and points represent the annual count of salamanders in CC sections. The green line and points show the annual count of salamanders in OG sections. Data source: Andrews Forest LTER Site._

------------------------------------

#### B. Salamander counts by channel classification and sections of Mack Creek in 2017.

_**Table 1**: Numbers of salamanders within Mack Creek in 2017. Values indicated are counts and proportions (by sections) in channel units (cascades, pool, and side-channel). In CC section, 67.1% observations are from cascades (n = 247), 8.4% from pools (n = 31), and 24.5% from side channels (n = 90). In OG section, 62.8% observations are from cascades (n = 201), 14.1% from pools (n = 45), and 23.1% from side channels (n = 74). Data source: Andrews Forest LTER Site._

```{r}
salamander_2017 <- salamander %>% 
  filter(year == 2017)

# create contigency table: count section and channel class, then transform to wider format
count_2017 <- salamander_2017 %>% 
  count(section, unittype) %>% 
  pivot_wider(names_from = unittype, values_from = n)

# add percentage of channel class in each sections to the contigency table
props_2017 <- count_2017 %>% 
  adorn_percentages(denominator = "row") %>% 
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns(position = "front")

# format the table 
props_2017 %>% 
  kable(col.names = c("Sections", "Cascades","Pool", "Side-channel")) %>% 
  kable_styling(bootstrap_options = "striped",
                position = "center",
                full_width = FALSE)
```

------------------------------------

#### C. Is there a significant difference in where in the channel Pacific giant salamanders are located between the two forest condition (OG and CC)?
```{r}
chi_2017 <- chisq.test(count_2017[, -1])
```

There is no significant effect of forest conditions (OG/CC) on where in the channel salamanders are located (cascades/pool/side-channel) ($\chi$^2^(`r chi_2017$parameter`) = `r round(chi_2017$statistic,2)`, *p* = `r round(chi_2017$p.value, 3)`). Salamanders from old growth and clear cut sections in the forest have no different chances to be spotted in cascades (OG = `r props_2017[[2, 2]]`, CC = `r  props_2017[[1, 2]]`), pool (OG = `r props_2017[[2, 3]]`, CC = `r props_2017[[1, 2]]`), and side-chanel (OG = `r props_2017[[2, 4]]`, CC = `r props_2017[[1, 4]]`).

------------------------------------

#### D. Compare weights of Pacific giant salamanders in OG and CC forest sections of the creek in 2017.
```{r, include = FALSE}
# check normality
ggplot(data = salamander_2017, aes(x = weight)) +
  geom_histogram(aes(fill = section)) +
  facet_wrap(~section)

ggplot(data = salamander_2017, aes(sample = weight)) +
  geom_qq(aes(fill = section)) +
  facet_wrap(~section)

# note: the data set is not normally distributed
```


```{r}
weight_summary <- salamander_2017 %>% 
  group_by(section) %>% 
  summarize(mean = round(mean(weight, na.rm = TRUE), 2),
            stdv = round(sd(weight, na.rm = TRUE), 2),
            count = n(),
            median = round(median(weight, na.rm = TRUE), 2)
            )

# the dataset has large numbers of data which allow using t-test on mean

# t-test on weight from the two forest sections
ttest_section_2017 <- t.test(weight ~ section, data = salamander_2017)

# effect size
weight_OG <- salamander_2017 %>% 
  filter(section == "OG")
weight_CC <- salamander_2017 %>% 
  filter(section == "CC")
d_section <- cohen.d(weight_OG$weight, weight_CC$weight, na.rm = TRUE)
```

Although weight in both forest sections (OG/CC) are positively skewed, the large number of data (OG = `r weight_summary$count[2]`, CC = `r weight_summary$count[1]`) allow us to use a two-sample t-test to compare the mean (Figure 2). 


```{r}
ggplot() +
  geom_beeswarm(data = salamander_2017, 
                aes(x = section, 
                    y = weight),
                size = 0.2) +
  geom_point(data = weight_summary, 
             aes(x = section, y = mean), 
             size = 1,
             color = "red") +
  geom_errorbar(data = weight_summary,
                aes(x = section, 
                    ymin = mean - stdv,
                    ymax = mean + stdv),
                width = 0.05,
                color = "red") +
  scale_x_discrete(labels = c("Clear cut", "Old growth")) +
  labs(x = "Forest section", y = "Salamanders weight (gram)") +
  theme_light()
```

_**Figure 2**: The weight of Pacific gaint salamanders in forest sections (CC/OG) in 2017. The distributions of weight in both forest sections are positively skewed. Salamanders in CC forest section have weight `r weight_summary$mean[1]` $\pm$ `r weight_summary$stdv[1]` gram (n = `r weight_summary$count[1]`) while those in OG sections have weight `r weight_summary$mean[2]` $\pm$ `r weight_summary$stdv[2]` gram (n = `r weight_summary$count[2]`). Data source: Andrews Forest LTER Site._


For 2017 observations, mean weight (gram) of Pacific gaint salamanders measured in samples collected from OG sections (`r round(ttest_section_2017$estimate[[2]], 2)` $\pm$ `r weight_summary$stdv[2]`, n = `r weight_summary$count[2]`) differed significantly from salamander mean weight in CC sections (`r round(ttest_section_2017$estimate[[1]], 2)` $\pm$ `r weight_summary$stdv[1]`, n = `r weight_summary$count[1]`) by a two-sample t-test (t(`r floor(ttest_section_2017$parameter)`) = `r round(ttest_section_2017$statistic, 2)`, *p* = `r ifelse(ttest_section_2017$p.value < 0.001, "< 0.001", round(ttest_section_2017$p.value, 3))`, 95% CI = (`r round(ttest_section_2017$conf.int, 2)`)). 

Based on *p*-value, there was a `r round(ttest_section_2017$p.value, 3)*100`% chance that we could have found two samples at least as different as ours if they were drawn from two populations with the same means, which is not signifcantly low. 

In addition, if we took many samples and found the confidence intervals, we'd expect 95% of those confidence intervals to included the population mean differece. Therefore, it's highly likely that a calculated confidence interval contains the population mean difference, where 0 was included. The difference in salamaders weight between OG and CC forest section was `r round(weight_summary$mean[1] - weight_summary$mean[2], 2)`mm, with a negligible effect size (Cohen's d = `r round(d_section$estimate, 2)`). 

Therefore, we failed to reject the null hypothesis that the weight of Pacific gaint salamanders measured in OG and CC forest sections were the same. The difference in forest conditions did not impact salamanders' weight in Mack Creak samples.

---------------------------------

#### E. Weights of Pacific gaint salamanders in pools, cascades, and side-channels of Mack Creek in 2017.

```{r}
# summary data of weight among channel classifications
weight_summary_unittype <- salamander_2017 %>% 
  group_by(unittype) %>% 
  summarize(mean = round(mean(weight, na.rm = TRUE), 2),
            stdv = round(sd(weight, na.rm = TRUE), 2),
            count = n(),
            se = round(sd(weight, na.rm = TRUE)/sqrt(n()), 2)
            )
```


```{r}
# visualize salamander weights among the three channel classifications

ggplot() +
  geom_beeswarm(data = salamander_2017, 
                aes(x = unittype, 
                    y = weight),
                size = 0.2) +
  geom_point(data = weight_summary_unittype, 
             aes(x = unittype, 
                 y = mean), 
             size = 1,
             color = "red") +
  geom_errorbar(data = weight_summary_unittype,
                aes(x = unittype, 
                    ymin = mean - stdv,
                    ymax = mean + stdv),
                width = 0.05,
                color = "red") +
  scale_x_discrete(labels = c("Cascades", "Pool", "Side-channel")) +
  labs(x = "Channel classificantion", y = "Salamanders weight (gram)") +
  theme_light()
```
_**Figure 3**: The weight of Pacific gaint salamanders in different channel locations (cascades/pool/side-channel) in 2017. The distributions of weight in all three locations are positively skewed. Salamanders in cascades have weight `r weight_summary_unittype$mean[1]` $\pm$ `r weight_summary_unittype$stdv[1]` gram (n = `r weight_summary_unittype$count[1]`, SE = `r weight_summary_unittype$se[1]`), those in the pools have weight `r weight_summary_unittype$mean[2]` $\pm$ `r weight_summary_unittype$stdv[2]` gram (n = `r weight_summary_unittype$count[2]`, SE = `r weight_summary_unittype$se[2]`), and those in the side-channels have weight `r weight_summary_unittype$mean[3]` $\pm$ `r weight_summary_unittype$stdv[3]` gram (n = `r weight_summary_unittype$count[3]`, SE = `r weight_summary_unittype$se[3]`). Data source: Andrews Forest LTER Site._


```{r, include = FALSE}
# check equal variance
unittype_levene <- leveneTest(weight ~ unittype, data = salamander_2017)
```
Although the weight of Pacific gaint salamanders is positively skewed, we can still conduct a one-way unbalanced ANOVA test because the variance are close (Levene's F(`r round(unittype_levene[[1,1]], 2)`) = `r round(unittype_levene[[1,2]], 2)`, *p* = `r round(unittype_levene[[1,3]], 3)`) and sample sizes are large (cascades = `r weight_summary_unittype$count[1]`, pool = `r weight_summary_unittype$count[2]`, side-channel = `r weight_summary_unittype$count[3]`) for the three groups. 

```{r}
unittype_aov <- aov(weight ~ unittype, data = salamander_2017)
aov_output <- unlist(summary(unittype_aov))

# pairwise comparison of group means
unittype_hsd <- TukeyHSD(unittype_aov)
```

Mean weights (gram) of Pacific gaint salamanders differed significantly across all channel classifications (one-way ANOVA with post-hoc Tukey's HSD, F(`r aov_output[1]`, `r aov_output[2]`) = `r round(aov_output[7],2)`, *p* = `r round(aov_output[9],3)`). This result means that salamanders spotted in at least one group of the cascades, pool, or side-channel have a different mean weight than the other groups. Pairwise comparison showed that mean weights of salamanders in side-channel and pool are significantly different (mean difference = `r round(unittype_hsd$unittype[3,1], 2)`, *p adj* = `r round(unittype_hsd$unittype[3,4], 3)`).


### Summary

### References